---
title: "Data624 Project 2"
author: "Johnny Rodriguez, Semyon Toybis, Adriana Medina"
date: "2024-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(readxl)
library(caret)
library(tidyverse)
library(VIM)
```

## Import Excel tables:

Below we import the excel data from Github

```{r}
stu_data_url <- 'https://github.com/amedina613/Data624-Project2/raw/refs/heads/main/StudentData.xlsx'

dest_file_stu_data <- 'StudentData.xlsx'

stu_eval_url <- 'https://github.com/amedina613/Data624-Project2/raw/refs/heads/main/StudentEvaluation.xlsx'
  
dest_file_stu_eval <- 'StudentEvaluation.xlsx'

download.file(stu_data_url, dest_file_stu_data, mode = 'wb')
download.file(stu_eval_url, dest_file_stu_eval, mode = 'wb')

stu_data_raw <- read_excel('StudentData.xlsx',col_names = T)
stu_eval_raw <- read_excel('StudentEvaluation.xlsx',col_names = T)

stu_data_raw <- as.data.frame(stu_data_raw)
stu_eval_raw <- as.data.frame(stu_eval_raw)
```

## Data Exploration:

First, we check the structured of the training set.

```{r}
str(stu_data_raw)
```

Next, we check the amount of missing values in each column.

```{r}
missing_values <- colSums(is.na(stu_data_raw))

print(missing_values)
```

## Data preparation

### kNN Imputation

The Brand Code column is a categorical variable. We will have to convert it to dummy variables to perform knn imputation.

First, we convert the column to a factor.

```{r}
stu_data_raw$`Brand Code` <- as.factor(stu_data_raw$`Brand Code`)
```

Next, we convert the Brand Code to dummy variables:

```{r}
dummy_variables <- dummyVars(~., data = stu_data_raw, fullRank = T)
training_data <- predict(dummy_variables, newdata = stu_data_raw)
```

```{r}
head(training_data)
```

We now have three dummy variable columns: Brand Code B, Brand Code C, and Brand Code D with zeros and ones. If all three columns have zero, then it means the Brand is A.

Next, we center and scale the data and perform KNN imputation.

```{r}
# Perform kNN imputation on columns (with scaling)
preProcess_model <- preProcess(training_data, method = c("center", "scale", "knnImpute"))
training_data_imputed <- predict(preProcess_model, training_data)
```

We now have zero missing values:

```{r}
colSums(is.na(training_data_imputed))
```

### Near Zero Variance predictors

Near zero variance predictors are identified and removed from the imputed numeric data set. These predictors have little variability and do not contribute meaningfully to analysis or modeling.

There is only one variable with near zero variance:

```{r}
colnames(training_data_imputed)[nearZeroVar(training_data_imputed)]
```

```{r}
# Remove near-zero variance predictors
training_data_imputed <- training_data_imputed[, -nearZeroVar(training_data_imputed)]

```

Our working training data-set is training_data_imputed

```{r}
head(training_data_imputed)
```
